### 2016-01-14 03-design.tex
Work on the design section

In this section, we describe how we parallelize each of the stages of the
algorithm's main loop. In most places, the usage of multithreaded parallelism
is straightforward; we will discuss details only where necessary. The
distributed design follows a master-worker paradigm, where in essence
the master controls the parallel operations and the workers perform the
calculations. For thread parallelism, we annotate the program with
OpenMP~\cite{OpenMP}. For distributed communication, we use MPI~\cite{MPI}.

@@

The largest data structures of the algorithm are $G$, $\pi$ and~$\phi$.
For the largest dataset in this paper, com-Friendster, $G$ has 1.8~billion
undirected edges. In our representation with directed edges, this takes
up 13.5GB. Our design lets $G$ reside only at the master. We observe that
the calculations in the update stages only require the subset of $G$ that
is accessed by the minibatch nodes, so the master scatters that subset to
the workers together with the scattering of the minibatch nodes. $\pi$~and
$\phi$ are float arrays of size $K \times N$. For our largest distributed
experiment, com-Friendster with $N$=64M and $K$=12K, each takes up~3TB. We
decided to store only $\pi$ and $\pi^{sum}$, and recalculate $\phi$ from these
whenever appropriate. $\pi$ is partitioned across the workers, and $\pi$
values are accessed via a DKV (distributed key-value) store. In both our
decisions to transport a subset of $G$ in each iteration and to recalculate
$\phi$, we gain a larger available memory at the cost of communication and
computation.

@@

{\Large\bf Discuss the lack of locality}

@@

Now, we describe the parallelization of each of the stages of the algorithm.

Exceptionally, the first stage, minibatch selection, is not itself
parallelized. However, in the distributed version its execution at the
master's is fully overlapped, using pipeline parallelism, while the workers are
performing \textit{update\_phi}. The minibatch is partitioned equally over the
workers; the relevant section of $G$ is distributed together with the
minibatch subsets.

After a worker has received its subset of the minibatch, it samples a
neighbor set for each of its minibatch nodes, using thread parallelism.

The next stage, \textit{update\_phi}, is the algorithm's dominant stage, not
only in calculation but also in memory accesses. It can be fully parallelized
because it is a data-parallel operation over each of the minibatch nodes.
\textit{update\_phi} first loads the $\pi$ values for its minibatch nodes and
their neighbors using the DKV store. The updates to $\phi$ for the minibatch
nodes are calculated independently using thread parallelism.

\textit{update\_pi} uses the value of $\pi/\phi$ for the minibatch nodes. The
update to $\pi$ can be done fully in parallel. Because of memory consistency,
this stage awaits completion of \textit{update\_phi} with an MPI barrier.
After the calculation, the updated values of $\pi/\pi^{sum}$ are written
through the DKV store.

\subsection{Pipelining of $\pi$ staging and computation}

@@

>>>>

This section describes how we parallelize the algorithm.
In most places, the usage of multithreaded parallelism
is straightforward; we will discuss details only where appropriate. The
distributed design follows a master-worker paradigm, where in essence
the master controls the parallel operations and the workers perform the
calculations. For thread parallelism, we annotate the program with
OpenMP~\cite{OpenMP}. For distributed communication, we use MPI~\cite{MPI}.

@@

The largest data structures of the algorithm are $G$, $H$, $\pi$ and~$\phi$.
For the largest dataset in this paper, com-Friendster, $G$ has 1.8~billion
undirected edges. In our representation with directed edges, this takes
up 13.5GB. Our design lets $G$ reside only at the master. We observe that
the calculations in the update stages only require the subset of $G$ that
is touched by the minibatch nodes, so the master scatters that subset to
the workers together with the scattering of the minibatch nodes. This way,
we trade a reduction in memory at the workers against the cost of limited
communication. In contrast, $H$ can be partitioned statically over all
machines for the parallel perplexity calculation.

$\pi$~and $\phi$ are 32-bit float arrays of size $K \times N$. For our
largest distributed experiment, com-Friendster with $N$=64M and $K$=12K, each
requires~3TB. We decided to store only $\pi$ and $\pi^{sum}$, and recalculate
$\phi$ from these whenever appropriate. Here, we trade a substantial reduction
in memory against limited computation. $\pi$ is partitioned across the workers,
and $\pi$ values are accessed via a DKV (distributed key-value) store.

@@

{\Large\bf NOTE: Discuss the lack of locality? Here?}

@@

This section describes the parallelization of each of the stages of the
algorithm's main loop. All but the first stage justify parallelisation for
high values of combinations of the  parameters $M$, $K$, $n$, and $|H|$.

The first stage, minibatch selection, is done at the master's, and it is
not itself parallelized. However, in the distributed implementation its
execution is overlapped using pipeline parallelism, while the workers are
performing \textit{update\_phi}. The minibatch is partitioned equally over
the workers; the relevant sections of $G$ are distributed together with the
minibatch subsets.

After a worker has received its subset of the minibatch, it samples a neighbor
set of size $n$ for each of its minibatch nodes, using thread parallelism.

The next stage, \textit{update\_phi}, is the algorithm's dominant stage, not
only in calculation but also in memory accesses. It performs $M\times n\times
K$ operations. It can be fully parallelized because it is a data-parallel
operation over each of the minibatch nodes. \textit{update\_phi} loads the
$\pi$ values for the local minibatch nodes and their neighbors from the
DKV store. The updates to $\phi$ for the minibatch nodes are calculated
independently using thread parallelism.

These updates are used by \textit{update\_pi} locally, besides the value of
$\pi/\phi$ for the minibatch nodes. The update to $\pi$ requires $M\times
K$ operations, and it is done in parallel over minibatch nodes. Because of
memory consistency, this stage awaits completion of \textit{update\_phi} with
an MPI barrier. After the calculation, the updated values of $\pi+\pi^{sum}$
are written through the DKV store.

\textit{update\_beta} requires $M\times K$ operations. It is preceded by an MPI
barrier to ensure that up-to-date $\pi$ values are read. \textit{update\_beta}
is split into four steps. The first stage partitions the minibatch across
machines to calculate contributions to $\nabla\theta$. The values of $\pi$ for
the local minibatch nodes are loaded from the DKV store. These calculations
are done with thread parallelism per minibatch node. In the second step,
a multithreaded summation is performed to calculate the contributions per
machine, followed by a distributed summation using an MPI reduce operation. The
third step calculates $\beta$ from $\theta$, sequentially at the master's; this
operation takes only $K$ steps. The resulting $\beta$ is broadcast with MPI
to the workers.

The calculation of perplexity requires $|H|\times K$ steps. Each of the
machines owns a subset of $H$. It loads up-to-date values of $\pi$ for each
node in its part of $H$, then for each of its edges in $H$ calculates the
contribution to the perplexity using thread parallelism. The resulting
contributions are summed to the global perplexity value in two stages,
with a local OpenMP reduction followed by a global MPI reduction.

Note that the calculation requirements per iteration do not depend
on $N$ or $|E|$, the size of the network graph. However, larger graphs are
expected to require more iterations to achieve convergence.

\subsection{Pipelining of computation and loading $\pi$}

As will be shown in Section~\ref{eval-pipeline}, loading $\pi$ is the
dominant contribution within the dominant stage \textit{update\_phi}. This
is the performance bottleneck of the distributed algorithm. We introduce
pipeline parallelism in order to shorten the critical path. Loading $\pi$
and calculating \textit{update\_pi} are partitioned into chunks, so the
calculation of a chunk is overlapped with loading the next chunk of $\pi$
values, and also with sampling the minibatch at the master's and deploying
it to the workers.

@@

Modern network technology allows access to the memory of remote machines
through the network cards (NICs) without involvement from the remote host
altogether~\cite{Infiniband,RDMA,Roce}. This has opened up the development of
extremely efficient remote memory services, often in the form of DKV stores.
This project stores $\pi$ in a DKV store in the collective memory of the
cluster, where $\pi[i]+\pi^{sum}[i]$ is the value for key $i$. We decided not
to use any of the existing implementations of RDMA DKV stores, and in stead
build our own DKV store on top of the ib-verbs API~\cite{ib-verbs}. The
main consideration is that our use case is unusually simple in a number
of important aspects. The KV layout is static in the sense that there are
no inserts or deletes after the initial population, which allows a static
partitioning of KV pairs over the machines. The values are all of the same size,
a vector of $K+1$ floats. The access pattern is very regular, because the
computation is partitioned into stages that are separated by barriers; a
stage either reads values or it updates values that it 'owns', so there
is no read/write or write/write contention. The existing DKV stores incur
overhead for insert/delete flexibility, concurrency control, or variable-sized
values. Our use case allows us to do any operation in exactly one RDMA read
or RDMA write. The Herd paper~\cite{Herd} argues that replacing RDMA reads by
an RPC to the server, followed by an unreliable write, is faster. However,
they show this only holds for small payload packets, up to 256B. A $\pi$
packet in our application is typically thousands to hundreds of thousands
of 4-byte floats.
